---
title: "MKT699: Empirical Methods in Marketing"
subtitle: "Lecture 3: Data Acquisition, Storage, and Provenance"
author: "Davide Proserpio"
institute: "Marshall School of Business"
date: "Fall 2026" #(updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      highlightSpans: true
      countIncrementalSlides: false
      ratio: "16:9"

---

```{css, echo=FALSE}
/* base sizing -- bumped up for classroom projection */
.remark-slide-content {
  font-size: 23px;
  padding: 1em 3em 1.5em 3em;
}
.remark-slide-content h1 { font-size: 46px; }
.remark-slide-content h2 { font-size: 38px; }
.remark-slide-content h3 { font-size: 30px; }

/* code -- wrap rather than overflow the slide */
.remark-code, .remark-inline-code {
  font-size: 20px;
}
.remark-slide-content pre {
  max-width: 100%;
  overflow-x: auto;
}
.remark-code-line {
  white-space: pre-wrap;
  word-break: break-word;
}

/* tables */
.remark-slide-content table {
  font-size: 21px;
}

.large2 { font-size: 165% }
.small90 { font-size: 92% }
.small75 { font-size: 82% }
```

```{r setup, include=FALSE}
library(fontawesome)
library(xaringan)
options(htmltools.dir.version = FALSE)
library(knitr)
opts_chunk$set(
  python.reticulate=FALSE,
  fig.align="center",
  dpi=300,
  cache=F
  )
```

# Where we left off

Last week was about verifying **code**. The model wrote something, it ran, and the
question was whether it did what you thought.

--

<br>

Today is about verifying **data**. Where did the rows come from, are they all there, and
can you prove it in eighteen months.

--

<br>

.large2[Same problem, one level down.]

---

# Table of contents

1. [When your data outgrows memory](#memory)

2. [SQL and relational databases](#sql)

3. [Collecting data: APIs](#apis)

4. [Scraping, honestly](#scraping)

5. [How LLMs changed data collection](#llm)

6. [Provenance and documentation](#provenance)

7. [Sharing data and code](#sharing)

---
class: inverse, center, middle
name: memory

# When your data outgrows memory

---

# The moment it happens

You have been working fine for two months. Then you get the full panel instead of the
sample, and R dies.

--

<br>

```
Error: cannot allocate vector of size 4.2 Gb
```

--

<br>

Almost every marketing PhD student hits this. What happens next determines a lot.

---

# The two bad responses

**Response 1: panic and buy nothing.** You spend three weeks writing chunked loops that
process the file 100k rows at a time, badly, with a bug in the last chunk.

--

<br>

**Response 2: subsample and never mention it.**

You take a 10% random sample "for development," write the whole paper on it, and never
go back. Or worse. You take the first 10 million rows, which is not a random sample,
it is whatever the file was sorted by.

--

<br>

.large2[Response 2 is the one that ends up in papers.]

---

# What actually fits

Rough arithmetic, and it is worth doing it before you start.

--

- A numeric (double) column: **8 bytes per row**
- An integer column: 4 bytes
- A character column: highly variable, often 50--100+ bytes per row
- A date: 8 bytes

--

<br>

10M rows × 30 numeric columns ≈ **2.4 GB**.

100M rows × 30 numeric columns ≈ **24 GB**.

---

# But that is the loaded size

You do not get to use all your RAM.

--

<br>

R copies. A `merge`, a `rbind`, a `mutate` on a data.frame, each one can duplicate the
object. `data.table` avoids some of this with `:=`, but not all of it.

--

<br>

**Working rule:** if the object is more than about **one third** of your RAM, you are
going to have a bad time.

--

<br>

16 GB laptop → keep it under ~5 GB. 64 GB workstation → under ~20 GB.

---

# Three real options

--

**(a) Put it in a database.** Query what you need, when you need it.

--

**(b) Use out-of-core tools.** `duckdb`, `arrow`. SQL and dplyr verbs that stream from
disk without loading everything.

--

**(c) Get a bigger machine.**

--

<br>

Option (c) is the one people skip. It is very often the right answer.

---

# In defense of just getting more RAM

Your time is worth more than the machine.

--

<br>

Three weeks of engineering a chunked pipeline, versus a machine with 256 GB of RAM where
the naive code just works.

--

<br>

You have access to CARC, the USC cluster we talked about in Week 1. Nodes there have
far more memory than your laptop, and it costs you nothing.

--

<br>

.large2[If your problem fits on a bigger machine, use a bigger machine.]

---
class: inverse, center, middle
name: sql

# SQL and relational databases

---

# Why this still matters

Three reasons, and none of them is "SQL is a nice language."

--

**1. Structure is enforced.** A column declared `INTEGER` cannot silently become
character because one row had a stray comma. Last week's entire Section 6.

--

**2. Data bigger than memory is normal.** The database was designed for it. Your laptop's
R session was not.

--

**3. One source of truth.** Your coauthor and you query the same table. Nobody is
emailing `listings_final_v3_USE_THIS.csv`.

---

# Core SQL, quickly

```{r, eval=FALSE}
SELECT state,
       room_type,
       COUNT(*)   AS n_listings,
       AVG(price) AS mean_price
FROM airbnb_sample
WHERE price > 0
GROUP BY state, room_type
HAVING COUNT(*) >= 50
ORDER BY n_listings DESC;

-- state room_type        n_listings  mean_price
-- NY    Entire home/apt       32543         233
-- CA    Entire home/apt       29719         251
-- NY    Private room          17973         103
```

--

<br>

That is most of what you will write. `SELECT`, `WHERE`, `GROUP BY`, `ORDER BY`.

--

I am not going to drill this. Ask a model, it will write correct SQL faster than I can
explain it.

---

# What I do want you to know cold

.large2[JOIN semantics.]

--

<br>

Not the syntax. The **semantics**, which rows survive, which rows duplicate, and which
rows disappear.

--

<br>

This is the same content as last week's silent merge errors, in a different language.

---

# The joins

```{r, eval=FALSE}
-- keeps only rows with a match on both sides
SELECT COUNT(*) FROM airbnb_sample l
INNER JOIN zip_market z ON l.zipcode = z.zip;
-- 199236

-- keeps ALL rows of listings; NULLs where no match
SELECT COUNT(*) FROM airbnb_sample l
LEFT JOIN zip_market z ON l.zipcode = z.zip;
-- 200000
```

--

<br>

`INNER` silently drops. `LEFT` preserves and marks.

--

**Default to `LEFT`.** Then count the NULLs. That count is information; an inner join
throws it away.

---

# The anti-join

```{r, eval=FALSE}
-- which listings have NO matching zip?
SELECT l.zipcode, l.city, l.state, COUNT(*) n
FROM airbnb_sample l
LEFT JOIN zip_market z ON l.zipcode = z.zip
WHERE z.zip IS NULL
GROUP BY l.zipcode, l.city, l.state
ORDER BY n DESC;

-- zipcode  city           state   n
-- NULL      New York       NY     93
-- NULL      Los Angeles    CA     45
-- NULL      San Francisco  CA     27
```

--

<br>

Or, more readably in a modern database:

```{r, eval=FALSE}
SELECT * FROM airbnb_sample
WHERE zipcode NOT IN (SELECT zip FROM zip_market);
```

---

# Why I am making a fuss about this

.large2[The anti-join is the single most underused diagnostic tool in empirical work.]

--

<br>

Run it on our data and the 764 unmatched rows explain themselves:

```{r, eval=FALSE}
unmatched[, .N, by = nchar(zipcode)][order(-N)]
#    nchar     N
# 1:    NA   649     <- missing
# 2:    10    71     <- ZIP+4
# 3:     6    25     <- malformed
```

--

<br>

Every join has a question attached to it: **what did not match?**

--

Almost nobody asks. They run the join, check that the row count is plausible, and move
on.

---

# Why I am making a fuss about this

Run it and the unmatched rows will show you one of three things: a **formatting** problem
(stripped leading zeros, whitespace, case), a **coverage** problem (your crosswalk ends in
2018 and your panel runs to 2021), or a **substantive pattern**: the unmatched are all
one city, one host type, one quarter.

--

That third case is the one that matters. Non-random attrition in a merge is a selection
problem wearing a technical costume.

---

# Make this a habit

```{r, eval=FALSE}
library(data.table)

merged <- merge(listings, zip_market, by.x = "zipcode", by.y = "zip",
                all.x = TRUE)

# always, immediately, no exceptions
unmatched <- listings[!zip_market, on = .(zipcode = zip)]

nrow(unmatched)                              # 764
unmatched[, .N, by = city][order(-N)][1:5]
unmatched[, .N, by = nchar(zipcode)]
```

--

<br>

Three lines. Do it after every single join in your pipeline.

--

If it returns zero rows, you have learned something. If it returns 400,000 rows, you have
learned a great deal more.

---

# Indexes, and why your query is slow

An index is a sorted lookup structure on a column. Without one, the database reads every
row.

--

<br>

```{r, eval=FALSE}
CREATE INDEX idx_airbnb_state_room ON airbnb_sample(state, room_type);
```

--

<br>

A query on `state` goes from scanning every row to jumping straight to the right block.

Seconds instead of minutes.

---

# But here is the thing

Most of you do not need MySQL or Postgres.

--

<br>

Setting up a server, managing users, loading data through an ETL step. That is real
overhead, and for a single-researcher project it buys you very little.

--

<br>

.large2[Use **DuckDB**.]

---

# DuckDB

An embedded analytical database. No server. No daemon. It is a library.

--

<br>

- Runs inside your R or Python process
- Queries **parquet and CSV files directly**, no import step
- Columnar and vectorized, built for analytical queries
- Streams from disk, so "bigger than RAM" is the normal case, not the exception

--

<br>

It is SQLite's shape with a column store's engine.

---

# What it looks like

```{r, eval=FALSE}
library(duckdb)
library(DBI)

con <- dbConnect(duckdb())

dbGetQuery(con, "
  SELECT state,
         room_type,
         COUNT(*)   AS n,
         AVG(price) AS mean_price
  FROM 'data/airbnb_parquet/listings_0*.parquet'
  WHERE price > 0
  GROUP BY state, room_type
  ORDER BY n DESC
")
#   state  room_type            n  mean_price
# 1    NY  Entire home/apt  32541       232.5
# 2    CA  Entire home/apt  29716       251.2
# 3    NY  Private room     17973       102.5
```

---

# What it looks like

There is no `CREATE TABLE`. There is no `LOAD DATA INFILE`. It reads the parquet files
where they sit.

--

<br>

Those four files are in the course repo. **Run this one.** Note the `*`. It reads all
four as if they were one table.

---

# The part that surprises people

That query works on **40 GB of parquet from a 16 GB laptop**.

--

<br>

DuckDB reads only the columns you asked for (`state`, `room_type`, `price`, not the other
23), only the row groups that can contain matching rows, and aggregates in a streaming
fashion.

--

<br>

Peak memory: a few hundred MB.

--

<br>

This is the single most useful practical thing in today's lecture.

---

# You can also skip SQL entirely

```{r, eval=FALSE}
library(duckdb); library(dplyr); library(dbplyr)

con <- dbConnect(duckdb())
listings <- tbl(con,
  "read_parquet('data/airbnb_parquet/listings_0*.parquet')")

listings |>
  filter(price > 0) |>
  group_by(state, room_type) |>
  summarise(n = n(), mean_price = mean(price, na.rm = TRUE)) |>
  collect()
```

--

<br>

`dbplyr` translates the verbs into SQL and pushes them into DuckDB. You write dplyr, the
database does the work.

--

`collect()` is where it runs and where the result comes back into R. Everything before it
is lazy.

---

# When you DO want a real server

- Multiple people writing to the same data concurrently
- Data that must live somewhere central with access control
- A production system, or a long-lived research infrastructure
- Your data provider hands you a Postgres connection string and that is that

--

<br>

Then: Postgres. Not MySQL, if you get a choice, Postgres has better analytical
functions and better type handling.

--

<br>

For an individual dissertation chapter, DuckDB over parquet is almost always right.

---

# Parquet vs CSV

.large2[Stop shipping CSVs.]

--

<br>

**CSV:** a text file with commas. It has no types. Every read is a guess.

--

**Parquet:** columnar binary format with a schema.

- Types are stored and preserved
- Compressed. Our sample: **28.5 MB as CSV, 7.2 MB as parquet**, 4× smaller, and it
  still knows `zipcode` is text
- Columnar: reading 3 of 80 columns reads ~3/80 of the bytes
- Row-group statistics let readers skip blocks entirely

---

# The types argument alone should convince you

Last week: leading zeros stripped from ZIP codes, dates read as strings, IDs read as
doubles losing precision, `"NA"` becoming a character level.

--

<br>

Every one of those is a **CSV** problem. Parquet stores that `zip` is `VARCHAR` and hands
it back as `VARCHAR`.

--

<br>

```{r, eval=FALSE}
library(arrow)
write_parquet(clean_panel, "data/clean/panel.parquet")
panel <- read_parquet("data/clean/panel.parquet")   # types intact
```

---
class: inverse, center, middle
name: apis

# Collecting data: APIs

---

# The hierarchy

--

1. **Someone already assembled it.** Nielsen, Kilts, Compustat, IRI, the library's
   subscriptions. Check first.

--

2. **There is an API**: an application programming interface, a documented way to
   request the data programmatically. Use it.

--

3. **There is no API.** Now we talk about scraping.

--

<br>

.large2[Most people skip step 1 and half of them skip step 2.]

---

# Why APIs first, always

- The data is **structured**. You get JSON, a machine-readable text format with
  labelled fields, not HTML you have to parse.
- It is **stable**. HTML changes weekly; APIs are versioned.
- It is **documented**. You know what each field means.
- It is **sanctioned**. You are not in a legal gray zone.
- It has **explicit rate limits**, so you know what you are allowed to do.

--

<br>

Scraping is what you do when none of that is available. It is not the default.

---

# Authentication

Nearly every API needs a key. That key is a credential.

--

<br>

```bash
# ~/.Renviron  -- NOT in your repo, NOT in git
CENSUS_API_KEY=abc123...
ANTHROPIC_API_KEY=sk-ant-...
YELP_API_KEY=...
```

```{r, eval=FALSE}
key <- Sys.getenv("CENSUS_API_KEY")
```

--

<br>

.large2[Week 1's rule, restated: **never commit a key**.]

--

Check your `.gitignore` right now. `.Renviron`, `.env`, `*_key.txt`.

---

# A basic request

```{r, eval=FALSE}
library(httr2)
library(jsonlite)

resp <- request("https://api.example.com/v1/listings") %>%
  req_headers(Authorization = paste("Bearer", Sys.getenv("API_KEY"))) %>%
  req_url_query(city = "los-angeles", limit = 50) %>%
  req_perform()

dat <- resp_body_json(resp)
```

--

<br>

That gets you 50 rows. You need 400,000. Everything interesting is in what comes next.

---

# Pagination

APIs never hand you everything at once.

```{r, eval=FALSE}
fetch_page <- function(offset) {
  request("https://api.example.com/v1/listings") %>%
    req_headers(Authorization = paste("Bearer", Sys.getenv("API_KEY"))) %>%
    req_url_query(city = "los-angeles", limit = 50, offset = offset) %>%
    req_perform() %>%
    resp_body_json()
}

all_pages <- list()
offset <- 0
repeat {
  page <- fetch_page(offset)
  if (length(page$results) == 0) break
  all_pages[[length(all_pages) + 1]] <- page$results
  offset <- offset + 50
}
```

--

Common schemes: offset/limit, page numbers, and **cursor tokens** (the response tells you
where to continue). Read the docs; guessing wastes a day.

---

# The pagination trap

Many APIs cap total results. You request 400,000 and the API quietly stops at 10,000.

--

<br>

Your loop terminates. You have data. Nothing errored.

--

<br>

**You now have a non-random 10,000-row sample**, ordered by whatever the API's default
sort is, usually relevance or recency.

--

<br>

Check the documented maximum. Then check whether your collected count is suspiciously
round.

---

# Rate limits and backoff

Every API has a rate limit. Exceed it and you get `429 Too Many Requests`. Exceed it
aggressively and you get banned.

```{r, eval=FALSE}
req %>% req_throttle(capacity = 10, fill_time_s = 60)   # 10 req/min
```

---

# Rate limits and backoff

And when a request fails transiently, do not retry immediately in a tight loop. That is
how you get banned instead of throttled.

--

```{r, eval=FALSE}
resp <- request(url) %>%
  req_retry(
    max_tries = 5,
    backoff = function(i) 2^i,          # 2, 4, 8, 16, 32 seconds
    is_transient = function(resp) resp_status(resp) %in% c(429, 500, 502, 503)
  ) %>%
  req_perform()
```

--

<br>

Note `is_transient`. A `404` is not transient, the resource does not exist. Retrying it
five times is pointless.

--

A `429` and a `503` are transient. Wait and try again.

---

# Cache every response to disk

.large2[Never request the same thing twice.]

--

<br>

Write the **raw response**, before parsing, before cleaning.

```{r, eval=FALSE}
dir.create("data/raw/api/listings", recursive = TRUE, showWarnings = FALSE)

save_raw <- function(id, resp) {
  writeLines(resp_body_string(resp),
             sprintf("data/raw/api/listings/%s.json", id))
}
```

---

# Cache every response to disk

Save-then-parse, never parse-then-save. If you keep only the parsed data frame, you have
baked your parsing assumptions into the only copy of the data you have.

--

<br>

API calls cost money or quota, the endpoint may disappear, and you *will* discover a
parsing bug three weeks in and need to reparse without recollecting.

This is Week 1's raw-data immutability rule, applied at the moment of collection.

---

# Idempotent collection

.large2[Your collection script must be interruptible and resumable.]

--

<br>

It will run for six hours. Your laptop will sleep, the VPN will drop, the API will have
an outage, or you will hit Ctrl-C by accident.

--

<br>

If restarting means starting over, you have a script you are afraid of. And a script you
are afraid of does not get rerun, which means it does not get fixed.

---

# The pattern

```{r, eval=FALSE}
collect_all <- function(ids, outdir = "data/raw/api/listings") {
  dir.create(outdir, recursive = TRUE, showWarnings = FALSE)

  # 1. what do I already have?
  have <- tools::file_path_sans_ext(list.files(outdir, pattern = "\\.json$"))

  # 2. what is still missing?
  todo <- setdiff(as.character(ids), have)
  message(sprintf("%d done, %d remaining", length(have), length(todo)))

  # 3. request only that
  for (id in todo) {
    resp <- try(fetch_one(id), silent = TRUE)
    if (inherits(resp, "try-error")) { log_failure(id); next }
    writeLines(resp, file.path(outdir, paste0(id, ".json")))
  }
}
```

--

Run it again and it picks up where it stopped. Run it after it finishes and it does
nothing.

---

# Log the collection itself

```{r, eval=FALSE}
log_entry <- data.frame(
  id          = id,
  url         = url,
  status      = resp_status(resp),
  n_returned  = length(dat$results),
  accessed_at = Sys.time(),
  api_version = "v1"
)
```

--

<br>

When a referee asks "when was this collected?" you have an answer with a timestamp
instead of "sometime last spring."

--

When your 2019 counts look wrong, you can check whether those were collected on the day
the API was returning garbage.

---
class: inverse, center, middle
name: scraping

# Scraping, honestly

---

# The disclaimer, up front

.large2[I am not a lawyer. This is not legal advice.]

--

<br>

I am going to describe the landscape as I understand it, because you need to know it
exists. Not so you can make a legal determination yourself.

--

<br>

**For anything sensitive, or anything you plan to publish: talk to me first, and if it
involves personal data, talk to USC's IRB and to the general counsel's office.**

--

Those conversations take a week. Doing them after collection takes a year.

---

# The four things that create exposure

--

**1. Terms of service.** The site's contract with users. A ToS violation is a contract
issue, not usually a criminal one. But it is real, and it can matter for publication.

--

**2. Computer Fraud and Abuse Act (CFAA).** The federal anti-hacking statute. Turns on
"unauthorized access."

--

**3. Load and disruption.** If you degrade someone's service, the analysis changes
regardless of what the ToS says.

--

**4. Personal data.** GDPR (EU) and CCPA (California). Separate legal regime entirely.

---

# On the CFAA

*hiQ Labs v. LinkedIn* is the case people cite.

--

<br>

Broadly: US courts in that line of cases have been **reluctant to treat scraping of
publicly accessible web pages as "unauthorized access" under the CFAA**. The Supreme
Court's decision in *Van Buren* (2021) also narrowed how broadly "exceeds authorized
access" is read.

--

<br>

That is the general direction. It is not a green light, and I am stating it at a level of
generality on purpose.

--

<br>

And it applies to **public** pages. It says nothing about anything behind a **login** or a
**paywall**, about **circumventing** CAPTCHAs or IP blocks, or about ignoring a
cease-and-desist after you have received one.

--

.large2[Logging in changes the analysis. Do not treat it as a detail.]

---

# Personal data is its own problem

CFAA questions and privacy questions are separate. Winning one does not settle the other.

--

<br>

If you are collecting data about **identifiable individuals**, names, reviews tied to
profiles, photos, locations, employment histories. You may be in scope for GDPR or CCPA
regardless of whether the pages were public.

--

<br>

You may also be doing **human subjects research**, which is an IRB question.

--

<br>

Public availability is not consent. Those are different concepts.

---

# The honest summary

Recent US case law has generally been more permissive about scraping publicly accessible
pages than the pre-2019 conventional wisdom assumed.

--

<br>

**And** ToS violations, server load, authentication barriers, and personal data are all
live sources of exposure that case law does not resolve.

--

<br>

**And** none of that is a determination anyone in this room is qualified to make about
your specific project.

--

<br>

.large2[Come talk to me before you collect, not after.]

---

# Being a good citizen

Separate from what you are permitted to do: what you should do.

--

**Rate limit yourself.** One request every 1--3 seconds is polite. Ten concurrent workers
hammering a small business's website is not.

--

**Identify your scraper.**

```{r, eval=FALSE}
req_headers(req,
  `User-Agent` = "USC-Marshall-research (proserpi@marshall.usc.edu)")
```

--

A real contact address means someone emails you before they block you.

--

**Read `robots.txt`.** Not legally binding, but a clear statement of the operator's
wishes. And disregarding it looks bad in exactly the situation where you need it to look
good.

--

**Cache aggressively**, never fetch a page twice, and **stop when asked**. If you get a
cease-and-desist, you stop and you come talk to me.

---

# The tools

**Static HTML**: the page's content is in the source.

```{r, eval=FALSE}
library(rvest)
page  <- read_html("https://example.com/listing/123")
price <- page %>% html_element(".price") %>% html_text2()
```

--

<br>

Python: `requests` + `BeautifulSoup` or `lxml`.

--

Test: view source. If the number you want is in there, you are in the easy case.

---

# JS-heavy pages

If the content is loaded by JavaScript after the page renders, `rvest` gets you an empty
shell.

--

<br>

**Option 1, find the underlying API.** Open your browser's network tab, reload, and look
at the XHR requests. Very often the page is calling a JSON endpoint that you can call
directly.

--

<br>

.large2[Do this before reaching for a browser. It works more often than you expect.]

--

It is faster, more stable, and gives you structured data instead of HTML.

--

<br>

**Option 2, drive a real browser.** Playwright (the current default), Selenium, or
`chromote` in R. They render, click, scroll, and extract.

--

Cost: 10--100× slower, far more fragile, and much heavier on the target server, which is
the citizenship point again.

---

# Mobile app traffic

Last resort. Some data only exists in a mobile app.

--

<br>

You can proxy the phone's traffic through your machine and inspect the API calls the app
makes. `mitmproxy` is the usual tool.

--

<br>

Two warnings. Technically: apps increasingly use certificate pinning specifically to
prevent this.

--

Legally: this is further from "reading a public web page" than anything else on this
list. **Do not do this without talking to me first.**

---

# Scrapers break

.large2[Every scraper you write will break. The only question is when you notice.]

--

<br>

A CSS class gets renamed. A field moves. A layout gets an A/B test. An anti-bot system
gets deployed.

--

<br>

For a one-shot collection this is annoying. For **panel data** it is fatal.

---

# The panel problem

Your project needs prices observed weekly for six months.

--

<br>

That is a scraper running unattended for 26 weeks. On week 9, the site changes its
markup.

--

<br>

Your scraper does not error. `html_element(".price")` returns `NA`. Your loop writes a
row with a missing price and continues.

--

<br>

You find out in month five that you have **17 weeks of NAs** and no way to recover them.
The pages are gone.

---

# Monitoring

Do not check the data at the end. Check it every run.

```{r, eval=FALSE}
run_checks <- function(dat) {
  stopifnot(nrow(dat) > 0.8 * expected_n)
  stopifnot(mean(is.na(dat$price)) < 0.05)
  stopifnot(all(dat$price > 0 & dat$price < 100000))
  stopifnot(n_distinct(dat$listing_id) > 0.9 * expected_ids)
  invisible(TRUE)
}
```

--

<br>

Assertions on the **shape** of the data, not the content. Row count, missingness rate,
range, cardinality.

---

# Alerting

A check that nobody reads is not a check.

```{r, eval=FALSE}
result <- try(run_checks(dat), silent = TRUE)

if (inherits(result, "try-error")) {
  send_email(
    to      = "you@usc.edu",
    subject = "SCRAPER FAILED: listings, run 2026-03-14",
    body    = as.character(result)
  )
}
```

--

<br>

Simplest version: a cron job that emails you the row count every morning. You will notice
a zero.

--

Set a calendar reminder to look at the data weekly. The technology is less important than
the habit.

---

# The reproducibility problem

This one is specific to scraping and it is worse than people think.

--

<br>

An API you can point to. A dataset you can cite. A scrape is a snapshot of a website that
**no longer exists**.

--

<br>

Six months later the pages have changed. A referee cannot verify your data. A replicator
cannot rebuild it.

--

<br>

.large2[Neither can you.]

---

# Archive the raw HTML

.large2[This is non-negotiable and almost nobody does it.]

--

<br>

```{r, eval=FALSE}
fetch_and_archive <- function(url, id, date_stamp) {
  html <- httr2::request(url) %>%
    httr2::req_perform() %>%
    httr2::resp_body_string()

  path <- sprintf("data/raw/html/%s", date_stamp)
  dir.create(path, recursive = TRUE, showWarnings = FALSE)

  # gzip: HTML compresses ~10x
  con <- gzfile(file.path(path, paste0(id, ".html.gz")), "w")
  writeLines(html, con); close(con)

  html
}
```

--

Two scripts: `01-collect.R` touches the network, `02-parse.R` never does.

---

# What the archive buys you

- Fix a parsing bug without recollecting anything
- Extract a field you did not think to extract in month one
- Hand a replicator the HTML and have them reproduce your parse exactly
- Show a referee the actual page your number came from

---

# "But that is a lot of disk"

400,000 pages × 200 KB, gzipped at roughly 10:1 ≈ **8 GB**.

--

<br>

Storage costs approximately nothing. Your data does not exist twice.

--

<br>

<br>

.large2[If you scrape and do not archive the raw pages, your data is not reproducible by
anyone, including you.]

--

I will ask about this when I grade your milestone.

---
class: inverse, center, middle
name: llm

# How LLMs changed data collection

---

# What is genuinely new

Not "the model writes the scraper faster." That is nice but it is last week's lecture.

--

<br>

.large2[The new capability is **extraction from unstructured sources at scale**.]

--

<br>

Sources that were previously unusable (not expensive, *unusable*) are now within reach
of a single PhD student.

---

# What was previously impossible

--

- 40,000 PDF regulatory filings, each formatted differently
- Scanned documents from before 1995
- HTML where each page's structure differs because 12,000 restaurants each built their
  own menu page
- Product descriptions written by 200,000 different sellers with no schema
- Numbers that exist only inside chart images in old industry reports
- Six thousand hours of earnings calls

--

<br>

The old answer was: hire RAs, or pick a different question.

--

Most people picked a different question. That is the important part, this changes which
questions get asked.

---

# The boundary I want to draw

**This lecture:** extraction. Getting a field out of a source when the field is *there*
and the problem is that it is formatted badly.

--

<br>

**Next week:** measurement. Constructing a research construct that does not exist
verbatim anywhere, and validating it.

--

<br>

Extracting `price = 145` from a listing page is extraction.

Deciding whether a review "expresses frustration with the host" is measurement.

--

<br>

The line is blurrier than that, and I will come back to it.

---

# Structured output

The mechanism that makes this work: you specify a schema, and you get typed fields back.

```{r, eval=FALSE}
schema <- list(
  type = "object",
  properties = list(
    price          = list(type = "number"),
    bedrooms       = list(type = "integer"),
    pets_allowed   = list(type = "boolean"),
    parking        = list(type = "string",
                          enum = c("garage", "street", "none", "unknown")),
    lease_months   = list(type = "integer")
  ),
  required = c("price", "bedrooms", "pets_allowed", "parking")
)
```

--

Not free text you regex afterwards. A JSON object with the types you asked for.

---

# Why the schema matters more than the prompt

--

**It forces you to define your variables.** You cannot write `enum = c(...)` without
deciding what the categories are. That is a research decision, and the schema makes you
make it explicitly.

--

**It makes output parseable.** No "sometimes it says 'Yes' and sometimes 'yes, pets are
allowed'."

--

**It is documentation.** The schema file *is* your data dictionary. Commit it.

--

<br>

Put it in `prompts/`, next to the prompt. Week 1's directory structure was not decorative.

---

# The `unknown` category

Notice `enum = c("garage", "street", "none", "unknown")`.

--

<br>

Always give the model an explicit "not stated" option.

--

<br>

Without one, it will pick the most plausible value when the information is absent. You
will get a complete column with no missingness, and the missingness will have been
**silently imputed by a language model**.

--

<br>

.large2[A field with no NAs from an unstructured source should worry you.]

---

# Example: product attributes

Marketplace listings. Free-text descriptions, no structured attributes.

```{r, eval=FALSE}
prompt <- "Extract product attributes from this listing description.
Return only fields explicitly stated. Use null when a field is not mentioned.
Do not infer from the product category.

Description: {description}"
```

--

<br>

Schema: `material`, `color`, `size_cm`, `warranty_months`, `country_of_origin`.

--

<br>

"Do not infer from the product category" is doing real work in that prompt. Without it,
every leather item gets `country_of_origin = "Italy"`.

---

# Example: contracts and filings

SEC filings, franchise agreements, supplier contracts.

--

<br>

Extract: effective date, counterparty names, contract value, termination clause presence,
exclusivity terms, renewal structure.

--

<br>

This used to be a law-student-army task. There are published finance and accounting
papers whose entire contribution was hand-coding a few thousand contracts.

--

<br>

That contribution is now cheap. Which means the bar moved: the contribution has to be the
question, not the coding.

---

# Example: charts in old reports

Industry reports from the 1990s. The underlying data is gone. There is a bar chart in a
PDF.

--

<br>

Multimodal models can read values off it.

--

<br>

Approximately. With error. **Systematically** with error, possibly.

--

<br>

This one needs validation more than any of the others, and "validate against the subset
where the report also printed a table" is exactly the design you would want.

---

# Entity resolution

The classic painful task, and one where LLMs are genuinely strong.

--

<br>

```
"Procter & Gamble Co."     vs  "P&G"
"Procter and Gamble"       vs  "PROCTER & GAMBLE CO"
"Nestlé S.A."              vs  "Nestle SA"
"Alphabet Inc. (Google)"   vs  "Google LLC"
```

--

<br>

Fuzzy string matching gets you 80% and then plateaus, because the remaining 20% requires
**knowing that P&G is Procter & Gamble**. That is world knowledge, not string distance.

---

# The right way to do entity resolution

Do not ask the model to match every pair. That is O(n²) and expensive.

--

<br>

**1. Block.** Cheap filter first, same country, same industry, first letter, TF-IDF
similarity above a threshold. Cuts the candidate set enormously.

--

**2. Match.** Send the surviving candidate pairs to the model.

--

**3. Ask for confidence.** Have it return a match probability, not just a binary.

--

**4. Review the middle.** Auto-accept high confidence, auto-reject low, and hand-check
the ambiguous band. That band is usually small.

---

# Now the caveat

.large2[Extraction is measurement.]

--

<br>

I have spent ten slides telling you this is a solved data-engineering problem. It is
mostly true and it is not entirely true.

---

# What an extracted field actually is

.large2[Every field you extract with a model is a **generated variable**.]

--

<br>

It has an error rate. That error rate is almost certainly **not random**.

--

<br>

If long descriptions get extracted more accurately than short ones, and long descriptions
come from bigger sellers, your extraction error is correlated with a variable that is in
your regression.

--

<br>

That is not noise. That is bias.

---

# The three questions

For every extracted field, you owe an answer to:

--

**1. What is the error rate?** Against hand-labeled ground truth, on a sample you drew
properly.

--

**2. Is the error correlated with anything?** With treatment, with time, with document
length, with source.

--

**3. What does it do to your estimate?** Attenuation, or something worse.

--

<br>

Next week is entirely about those three questions. I am not answering them today.

--

<br>

A verbatim field (a price, a date, a name) is closer to parsing, and a few hundred
hand-checks will do. A field requiring judgment is measurement wearing extraction's
clothes.

--

.large2[When in doubt, treat it as measurement.]

---

# Cost control

You will run this over hundreds of thousands of documents. Costs compound.

--

**1. Cache by content hash.**

```{r, eval=FALSE}
key <- digest::digest(paste(prompt_version, model, document))
if (file.exists(cache_path(key))) return(readRDS(cache_path(key)))
```

--

Note `prompt_version` and `model` in the hash. Change the prompt, the cache correctly
misses. Do not hash only the document.

--

**2. Batch.** Most providers offer a batch API at a large discount for non-time-sensitive
jobs. Your annotation run is not time-sensitive.

--

**3. Tier your models.** Cheap model for easy fields, expensive model for hard ones.

--

**4. Escalate on disagreement.** Run the cheap model twice at temperature > 0. Where the
runs agree, keep it; where they disagree, escalate to the expensive model. Disagreement is
a decent proxy for difficulty, and it concentrates your spend on the hard cases.

---

# Test on 100 first

.large2[Never launch a 400,000-document job you have not run on 100.]

--

<br>

Run 100. **Read the output yourself.** All hundred.

--

<br>

You will find: a field that is always null because the schema name does not match the
prompt, a category the model never uses, an edge case you did not anticipate, or a
systematic misreading.

--

<br>

Fixing that after 100 documents costs a coffee. After 400,000 it costs your API budget
and a week.

---
class: inverse, center, middle
name: provenance

# Provenance and documentation

---

# The question

.large2[What does "the data" mean when part of it was generated by a model?]

--

<br>

Your panel has 40 columns. Six came from an administrative source. Twenty were derived by
your own code. Nine were scraped. Five were extracted by an LLM.

--

<br>

They are all just columns in a data frame. They look identical.

--

<br>

They have wildly different epistemic status, and nothing in the file records that.

---

# Provenance

For every variable in your analysis file: **where did it come from?**

--

<br>

Categories that matter:

- **Administrative / vendor**, someone else collected it, you trust their process
- **Survey / experimental**. You collected it under a design
- **API**, machine-collected, structured at source
- **Scraped**, machine-collected, parsed by you
- **LLM-extracted**, generated by a model
- **Derived**, computed by your code from other variables

--

<br>

A referee reading your table should be able to tell which is which. Usually they cannot.

---

# The provenance table

Maintain one file. One row per variable. Commit it.

.small75[

| variable | source | date_accessed | process | validation |
|:--|:--|:--|:--|:--|
| `listing_id` | Airbnb scrape | 2026-02-01 | `01-collect.R` | key, unique |
| `price` | Airbnb scrape | 2026-02-01 | `02-parse.R` CSS `.price` | range checked |
| `median_income` | ACS 5-yr 2019 | 2026-01-15 | Census API | vendor |
| `is_prof_host` | derived |, | `03-clean.R:L84` |, |
| `has_pool` | LLM extract | 2026-02-20 | `prompts/amenities_v3.md`, `claude-sonnet-4-5` | 500 labeled, κ=0.89 |
| `complaint_type` | LLM extract | 2026-02-22 | `prompts/complaints_v2.md`, `claude-opus-4-1` | **pending** |

]

--

That last row is honest and it is fine at this stage. An empty validation column that
stays empty is not fine.

---

# Why the model string matters

`claude-sonnet-4-5` is not a permanent artifact. Week 1: you cannot `renv::snapshot()` a
model.

--

<br>

Recording the exact model string and the date is the closest thing to a version pin you
have.

--

<br>

It lets a future reader say "this was extracted with a 2026 model". Which is information
they need and which is otherwise unrecoverable.

--

<br>

And it lets *you*, in 2028, re-extract with a new model and check whether your
conclusions moved.

---

# Raw data immutability, again

.large2[`data/raw/` is read-only. Third time I have said this.]

--

<br>

This week it means something specific:

- The API's JSON, exactly as returned
- The gzipped HTML, exactly as fetched
- The raw model responses, **before** you parsed them into fields

--

<br>

That last one is the one people skip.

---

# Save the raw model output

You asked for JSON. You parsed it into five columns. You threw away the response.

--

<br>

Then you discover the model was also returning a `reasoning` field you ignored, which
would have told you which cases it was unsure about.

--

<br>

Or you discover a parsing bug in how you handled `null`.

--

<br>

Re-running costs money and, because the model may have changed, **will not give you the
same output**. Re-parsing costs nothing.

---

# The sample construction table

The most useful table in your appendix, and it costs you twenty minutes.

.small90[

| Step | Rows | Units | Dropped | Reason |
|:--|--:|--:|--:|:--|
| Raw scrape | 4,201,338 | 218,442 |, |, |
| Parsed successfully | 4,118,902 | 214,003 | 82,436 | page structure changed wk 9--12 |
| Non-missing price | 3,904,551 | 209,881 | 214,351 | price field absent |
| Matched to tract | 3,655,120 | 196,214 | 249,431 | geocode failure |
| Active ≥ 6 months | 2,890,447 | 121,338 | 764,673 | panel balance requirement |
| **Analysis sample** | **2,890,447** | **121,338** | | |

]

--

<br>

Row 2 tells you your scraper broke in weeks 9--12. Row 4 tells you geocoding dropped 6%
of units, and you should check whether those were random.

---
class: inverse, center, middle
name: sharing

# Sharing data and code

---

# The policy landscape

Most top marketing and economics journals now require code, and where possible data, as a
condition of acceptance.

--

<br>

*Marketing Science*, *Management Science*, *JMR*, the AEA journals, *Econometrica*, all
have data and code availability policies. Several have a dedicated replication editor who
runs your package before publication.

--

<br>

Check your target journal's current policy. They have tightened over the past few years
and they will keep tightening.

---

# Proprietary data

The obvious problem: Nielsen, Compustat, and your platform partner's data cannot be
posted.

--

<br>

Journals know this. The standard accommodations:

--

- **Post all code**, unconditionally. Code is essentially never proprietary.
- **A data availability statement**: what the data is, who owns it, how someone else
  could obtain it, what it cost, whether there is an NDA.
- **Synthetic or simulated data** with the same schema, so a replicator can run the
  pipeline end to end and verify it executes.
- **Aggregated or de-identified extracts** where the agreement permits.

---

# The synthetic data trick

Generate a fake dataset with the same column names, types, and rough distributions.

```{r, eval=FALSE}
# code/00-make-synthetic.R
set.seed(699)
synth <- data.frame(
  listing_id = sprintf("L%06d", 1:50000),
  city_id    = sample(1:40, 50000, replace = TRUE),
  year       = sample(2015:2019, 50000, replace = TRUE),
  price      = round(rlnorm(50000, 4.8, 0.6), 2),
  has_pool   = rbinom(50000, 1, 0.22)
)
arrow::write_parquet(synth, "data/synthetic/listings.parquet")
```

--

<br>

A replicator runs your entire pipeline on it. The coefficients are meaningless. **The
code either runs or it does not**, and that is what they are checking.

---

# Where to put it

- **openICPSR**: the AEA's repository. If you are targeting an econ journal, this is
  where the package goes.
- **Harvard Dataverse**, general purpose, widely used in social science.
- **Zenodo**, general purpose, integrates directly with GitHub, gives you a DOI on
  release.

--

<br>

All three issue a **DOI**. That is the point.

--

<br>

.large2[A GitHub link is not an archive.]

You can force-push over it, delete it, or make it private. Referees know this.

---

# Privacy and licensing

Two constraints that are easy to forget until they block you.

--

<br>

**You may not be allowed to redistribute what you were allowed to collect.**

--

Scraped content is generally still under someone's copyright or ToS. Collecting it for
research analysis and republishing it as a dataset are different acts.

--

<br>

Post the **derived variables** you need, and the **code** that produced them. Not the
scraped corpus.

--

<br>

**And de-identification helps without being a guarantee.** Removing names does not
anonymize a dataset, quasi-identifiers in combination (ZIP, birth date, gender; or a
review history) can be enough to re-identify.

--

That is an **IRB** conversation, and IRB timelines are measured in weeks. Have it early.

---

# What a good replication package looks like

```
replication/
├── README.md
├── DATA_AVAILABILITY.md
├── code/
│   ├── 00-run-all.R          <- master script
│   ├── 01-collect.R          <- MARKED: requires API key
│   ├── 02-parse.R
│   ├── 03-extract-llm.R      <- MARKED: costs ~$340
│   ├── 04-validate.R
│   └── 05-analysis.R
├── data/
│   ├── synthetic/
│   └── intermediate/         <- cached outputs of the marked steps
├── prompts/
├── output/
└── renv.lock
```

---

# The master script

```{r, eval=FALSE}
# 00-run-all.R
source("code/02-parse.R")      # ~4 min
source("code/04-validate.R")   # ~1 min
source("code/05-analysis.R")   # ~12 min, produces all tables/figures

# STEPS NOT RUN BY DEFAULT
# ------------------------
# code/01-collect.R   Requires Airbnb access + ~6 hours.
#                     Output cached: data/intermediate/raw_listings.parquet
#
# code/03-extract-llm.R  Requires ANTHROPIC_API_KEY. ~$340, ~9 hours.
#                     Output cached: data/intermediate/extracted.parquet
```

--

One command reproduces every table and figure. The steps a replicator cannot run are
**named, priced, and their outputs shipped**.

---
class: inverse, center, middle

# Wrapping up

---

# What I want you to take from today

--

**Storage:** DuckDB over parquet. Get a bigger machine before you get clever.

--

**Collection:** APIs before scraping. Idempotent, cached, logged.

--

**Scraping:** archive the raw HTML. Talk to me before you collect anything sensitive.

--

**Extraction:** genuinely transformative, and every extracted field is a generated
variable.

--

**Provenance:** one row per variable, source, date, process, validation status.

--

<br>

.large2[If you cannot say where a column came from, you do not have data. You have
numbers.]

---

# Assignment 1 — due next week

**Build a reproducible data pipeline.**

--

Given a raw source, produce a clean analysis file, with:

- A collection script that is **idempotent** and logs what it got
- Raw data preserved unmodified
- Anti-join diagnostics after every merge, reported
- A sample-construction table
- A provenance table covering every variable
- A `README` such that I can clone it and run it

--

<br>

I will clone it and run it. If it does not run from a clean clone, it does not run.

---

# Week 3 milestone

Due before next class, Sep 16. Every milestone from here on works this way: assigned in
class, due before the following one. The syllabus table lists both dates.

--

<br>

**Data acquired, pipeline reproducible.**

--

<br>

Not "I found a website." Not "I have a plan."

--

<br>

Data on disk, in `data/raw/`, immutable, with a script that turns it into something you
can analyze.

--

<br>

<br>

If your data source turned out to be infeasible, that is a legitimate outcome. But I
need to know **this week**, not in Week 6, because we will need to change the project.

---

# Next week

**LLMs as measurement instruments I: validation.**

--

<br>

You will have the data. Today was about getting it and being able to prove where it came
from.

--

<br>

Next week we ask a harder question: does your variable **measure what you say it
measures**?

--

<br>

Human labels, gold standards, agreement statistics, sampling for validation, and what
happens to your standard errors when the dependent variable was generated by a model.

---
class: inverse, center, middle

# Questions?

### Bring a laptop. We are setting up DuckDB and running your first collection script.
