---
title: "MKT699: Empirical Methods in Marketing"
subtitle: "Lecture 4: LLMs as Measurement Instruments I — Promise and Validation"
author: "Davide Proserpio"
institute: "Marshall School of Business"
date: "Fall 2026" #(updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      highlightSpans: true
      countIncrementalSlides: false
      ratio: "16:9"

---

```{css, echo=FALSE}
/* base sizing -- bumped up for classroom projection */
.remark-slide-content {
  font-size: 23px;
  padding: 1em 3em 1.5em 3em;
}
.remark-slide-content h1 { font-size: 46px; }
.remark-slide-content h2 { font-size: 38px; }
.remark-slide-content h3 { font-size: 30px; }

/* code -- wrap rather than overflow the slide */
.remark-code, .remark-inline-code {
  font-size: 20px;
}
.remark-slide-content pre {
  max-width: 100%;
  overflow-x: auto;
}
.remark-code-line {
  white-space: pre-wrap;
  word-break: break-word;
}

/* tables */
.remark-slide-content table {
  font-size: 21px;
}

.codesm .remark-code, .codesm .remark-code-line { font-size: 16px; line-height: 1.25; }
.large2 { font-size: 165% }
.small90 { font-size: 92% }
.small75 { font-size: 82% }
```

```{r setup, include=FALSE}
library(fontawesome)
library(xaringan)
options(htmltools.dir.version = FALSE)
library(knitr)
opts_chunk$set(
  python.reticulate=FALSE,
  fig.align="center",
  dpi=300,
  cache=F
  )
```

# Where we are

Three weeks of infrastructure: repositories, verification, data provenance.

--

<br>

Today we start on measurement, which is what the project is built around.

---

# Table of contents

1. [The measurement problem](#problem)

2. [How we got here](#history)

3. [The promise](#promise)

4. [The annotation pipeline](#pipeline)

5. [Designing a validation study](#validation)

6. [Prompt sensitivity](#sensitivity)

7. [Where this leaves you](#close)

---
class: inverse, center, middle
name: problem

# The measurement problem

---

# What marketing actually studies

Make a list of the constructs central to our field.

--

- Brand positioning
- Service quality
- Consumer intent
- Product similarity
- Complaint type
- Emotional tone
- Visual style

--

<br>

Now ask: which of these has a column in a dataset?

---

# None of them

The constructs we care about live in **unstructured data**.

--

<br>

Text, images, audio, video. Reviews, descriptions, call transcripts, ad creative, listing
photos.

--

<br>

And for most of the history of the field, we could not measure them at scale.

---

# What we did instead

**(a) Hand-code a small sample.**

Hire two RAs. Write a codebook. Have them label 800 documents. Check that they agree
with each other. Six weeks and $4,000 later you have 800 observations.

--

**(b) Use a crude proxy.**

Star ratings for quality. Keyword counts for topic. A sentiment dictionary for tone. Word
count for effort.

--

**(c) Do not study it.**

The most common option.

---

# Option (c) is the expensive one

It leaves no trace, because papers that were never written are not in the literature.

--

<br>

But it shaped the field. Entire classes of question were off the table, not because they
were uninteresting, but because there was no instrument.

--

<br>

The set of questions a field asks is determined by what it can measure.

---

# What changed

LLMs give you a fourth option: measure the construct directly, on the full corpus, at a
small fraction of what RAs cost.

--

<br>

Gilardi et al. put the cost ratio at roughly **1/30** of crowd work. We'll get to that
paper in a moment.

--

<br>

A pipeline that used to cost $4,000 and six weeks now costs about $130 and an afternoon,
and runs on 200,000 documents instead of 800.

---

# Why this is a research opportunity

Constructs that were previously unmeasurable at scale are now measurable at
scale.

--

<br>

That is an expansion of the set of **answerable questions**, not only a cheaper way to
answer old ones.

--

<br>

Which is where dissertation topics come from.

---

# How to think about this for your project

Do not start from "what can I do with an LLM?"

--

<br>

Start from: what did I want to know, and stop pursuing, because I could not measure it?

--

<br>

Go back to an idea you abandoned and ask whether the measurement constraint still binds.

---

# A worked instance

"Do firms that describe their products in more *aspirational* language charge higher
prices?"

--

<br>

In 2015: you need a codebook, four RAs, and a year. Almost nobody did.

--

In 2020: you build a dictionary of aspirational words. A referee points out, correctly,
that the dictionary is arbitrary.

--

<br>

In 2026: you write a prompt, run it on 300,000 product descriptions, and validate it.

--

<br>

The paper is now feasible. Whether it is *credible* is the rest of this course.

---

# The catch

All of that is true, and it comes with a cost.

--

<br>

A cheap instrument is still an instrument. It has error, that error has structure, and
the structure matters.

--

<br>

This lecture: how to build one and check it.

Next lecture: why checking it is not enough.

---

# Vocabulary for today

Six words I will use constantly. Say so if I use one I have not defined.

--

.small90[
| term | what it means |
|:--|:--|
| **instrument** | the prompt plus the model: the procedure that turns a document into a label. Not an instrumental variable; that is Week 6. |
| **corpus** | the body of documents you are measuring. Your 200k reviews. |
| **annotation** | assigning a label to a document. Coding, by another name. |
| **gold standard** | the labels you trust, normally human, always small. |
| **zero-shot** | asking a model to do a task with no worked examples. |
| **downstream** | whatever you do with the variable afterwards, the regression. |
]

--

<br>

None of this is deep, but the papers assume it.

---
class: inverse, center, middle
name: history

# How we got here

---

# Twenty years in five slides

This is not a deep learning lecture. You do not need one to do this well.

--

<br>

You need enough of the history to know **which tool fits which task**, because using the
expensive tool where a cheap one works is a common mistake.

--

<br>

.small90[Background reading, if you have no NLP foundation: Hartmann & Netzer (2023),
"Natural Language Processing in Marketing," in *Artificial Intelligence in Marketing*
(Review of Marketing Research, Vol. 20), Emerald.]

---

# Stage 1: bag of words

Represent a document as counts of the words it contains. Order discarded.

--

<br>

**Could do:** topic proportions, dictionary methods, simple classification with enough
training data.

--

**Could not do:** anything requiring word order, negation, or context.

--

<br>

"Not good" and "good" are nearly identical documents. That is the limitation in one
example.

---

# Stage 2: TF-IDF

Down-weight words that appear everywhere, up-weight words distinctive to a document.

--

<br>

Still bag of words. Still no order. But now "the" stops dominating your similarity
measure.

--

<br>

.small90[Cheap, interpretable, and still a reasonable baseline for document retrieval.]

---

# Stage 3: word embeddings

word2vec, GloVe. Represent each word as a list of numbers, a **vector**, learned from
which words it tends to appear near.

--

<br>

.small90[Concretely: "excellent" becomes something like `[0.21, -0.88, 0.04, ...]`, a few
hundred numbers long. The numbers mean nothing individually. What matters is that two
words used similarly end up with similar lists.]

--

<br>

Words used in similar contexts get similar vectors. "Excellent" lands near "superb."

--

<br>

**Could do:** semantic similarity, analogy, clustering, transfer across corpora.

--

**Could not do:** disambiguate. "Bank" gets one vector. River bank, savings bank, same
vector.

---

# Stage 4: contextual embeddings

BERT and relatives. The vector for a word depends on the sentence it appears in.

--

<br>

"Bank" in *river bank* and "bank" in *bank account* now get different representations.

--

<br>

**Could do:** the best classification available at the time, given labeled training
data to **fine-tune** on (continue training the model on your own labeled examples so it
specializes)
on.

--

**Could not do:** work without that labeled training data. You still needed thousands of
examples.

---

# Stage 5: generative LLMs

The thing that changed is not accuracy. It is that **you no longer need training data**.

--

<br>

You write the construct definition in English. The model applies it.

--

<br>

The instrument is now a paragraph of prose instead of 5,000 labeled examples.

--

<br>

The barrier that fell was the *labeling* barrier, which is why this is a methods topic
rather than a computer-science one.

---

# Embeddings still have a job

Embeddings did not go away. They are still the right tool for a large class of problems.

--

<br>

**Use embeddings for:** similarity, nearest neighbors, clustering, deduplication,
retrieval, "which of these 50,000 products are close substitutes."

--

**Use a generative LLM for:** applying a construct definition, extraction, classification
against a codebook, anything requiring judgment.

---

# Why this matters for your budget

An embedding call is roughly two to three orders of magnitude cheaper than a generative
call on the same document.

--

<br>

If your task is "find similar product descriptions," and you loop 300,000 documents
through a generative model asking it to rate similarity, you have spent a thousand
dollars to do worse than a cosine similarity that costs eight.

--

<br>

Do not use a generative model where an embedding would do.

---
class: inverse, center, middle
name: promise

# The promise

---

# The paper that started it in social science

Gilardi, Alizadeh & Kubli (2023), "ChatGPT Outperforms Crowd Workers for Text-Annotation
Tasks," *PNAS* 120(30):e2305016120.

--

<br>

$n$ = 6,183 tweets and news articles. Several annotation tasks: relevance, stance, topic,
frame.

--

<br>

Compared against trained human annotators as the gold standard, with crowd workers
(MTurk) as the incumbent.

---

# The result

**Zero-shot**, no examples, no fine-tuning, just a prompt.

--

<br>

Accuracy exceeds crowd workers by roughly **25 percentage points** across tasks.

--

<br>

Intercoder agreement is also higher than among crowd workers.

--

<br>

At about **1/30** the cost.

---

# Why the result spread, and what got generalized

**What the paper showed.** On four annotation tasks on tweets, relevance, stance, topic,
and frame, a zero-shot prompt was more accurate than crowd workers, agreed with itself
more than they did, and cost about a thirtieth as much.

--

<br>

**What got generalized.** Crowd annotation was the standard way to scale text coding past
a few hundred documents, so the reading became: a zero-shot LLM is the best available
text classifier, for any task, and human labels and fine-tuned models can be skipped.

--

<br>

**Why that does not follow.** The comparison was against crowd workers, a low bar, on a
handful of judgment tasks. No fine-tuned classifier was in the race, and no task with a
fixed target and plenty of labels was tested.

--

<br>

The result is real. The generalization people drew from it is not.

---

# The corrective

Ziems, Held, Shaikh, Chen, Zhang & Yang (2024), "Can Large Language Models Transform
Computational Social Science?" *Computational Linguistics* 50(1):237–291.

--

<br>

13 models. 25 representative benchmark tasks. A systematic evaluation instead of a
handful of tasks.

--

<br>

If you read one paper from this section, read this one.

---

# What they find

LLMs do **not** beat fine-tuned models on classification.

--

<br>

LLMs do **well** at free-form generation, explanation, open coding, and summarizing
the reasoning behind a label.

--

<br>

Two findings in opposite directions, in the same paper.

---

# The point

"LLMs are good at annotation" is not a fact about LLMs. It is a claim about a **task**,
true for some tasks and false for others.

--

.small90[
| Fine-tune a small model when | Use an LLM when |
|:--|:--|
| thousands of labeled examples exist and the target is fixed | no labels exist and the label takes judgment |
| spam or not, from 50,000 flagged emails | does this pitch disclose failed prototypes? |
| product category from the description, with the catalog as labels | is this listing written in a professional voice? |
| sorting support tickets into twelve known categories | is this new app a copy of one of these five? |
| sentiment of reviews, with the star ratings as labels | is the complaint about the product or the shipping? |
]

---

# Why the split

**Left column.** The labels teach the model the exact target, so a fine-tuned small model
is cheaper, faster, and usually more accurate than a prompt.

--

<br>

**Right column.** No labels exist, so the LLM is the only instrument there is. That is
what makes it usable, and it is also why it has to be validated: nothing else has checked
it.

--

<br>

If you have a few hundred labels, they are worth more as a validation sample than as
training data. That is the design we build in the second half of today.

---

# Which case are you in?

Most of your projects are in the second case.

--

<br>

But you should be able to say *why*, in one sentence, in your paper.

--

<br>

"We use an LLM rather than a fine-tuned classifier because our construct has no existing
labeled corpus and requires applying a definition rather than pattern-matching a fixed
category."

--

<br>

A sentence like that belongs in your measurement section.

---

# The marketing paper

Li, Castelo, Katona & Sarvary (2024), "Frontiers: Determining the Validity of Large
Language Models for Automated Perceptual Analysis," *Marketing Science* 43(2):254–266.

--

<br>

Perceptual maps, the classic marketing object. Brands positioned in attribute space,
traditionally built from consumer surveys.

--

<br>

Can an LLM produce the map instead?

---

# What they find

Above **75% agreement** between human-generated and LLM-generated perceptual maps
**for certain categories**.

--

<br>

The qualifier carries the result.

--

<br>

Performance depends on the category: how much the model has seen about it, how
well-established the brands are, how conventional the attribute space is.

---

# Category-conditional validity

The instrument is valid **for some categories and not others**, and you cannot
tell which from the model's confidence.

--

<br>

This pattern recurs.

--

<br>

The question is not whether the LLM works but where. It works on the part of the world
it has seen a lot of, and degrades on the rest with no change in fluency.

---

# Assembling the three papers

.small90[
| Paper | Claim |
|:--|:--|
| Gilardi et al. (2023) | Beats crowd workers, 1/30 cost |
| Ziems et al. (2024) | Loses to fine-tuned models on classification |
| Li et al. (2024) | Valid for some categories, not others |
]

--

<br>

They are not contradictory. Each describes a different task and domain.

--

<br>

Validity is task-conditional, domain-conditional, and it must be **measured in
your setting**.

--

Another paper's validation does not transfer to your setting.

---
class: inverse, center, middle
name: pipeline

# The annotation pipeline

---

# What we're building

You have $n$ documents. You want a variable.

--

<br>

Five pieces:

1. A prompt that encodes your construct
2. A structured output format you can parse
3. A calling strategy that controls cost
4. Error handling for the calls that fail
5. Storage that keeps everything

--

<br>

None of this is deep, but each piece costs time if you get it wrong.

---

# Prompt design: the wrong mental model

Most people write annotation prompts like they are chatting.

--

<br>

> "Read this review and tell me if the customer is complaining about product quality."

--

<br>

It will produce output, but inconsistent output, and you will not be able to say what the
variable means.

---

# The right mental model

Your prompt is a **codebook**.

--

<br>

Everything you would have to write for a human research assistant, you write for the model. Same content,
same specificity.

--

<br>

Everything that applies to a codebook applies here, unchanged. The coder is a
model; the instructions are the same instructions.

---


# What a codebook contains

**1. The construct definition.** One or two sentences. Precise.

--

**2. Inclusion criteria.** What counts.

--

**3. Exclusion criteria.** What does not count, especially near-misses.

--

**4. Edge cases with resolutions.** The hard ones, decided in advance.

--

**5. A forced output format.** So there is exactly one way to answer.

---

# Edge cases

You will discover that your construct is ambiguous only by hitting the edges.

--

<br>

For "complaint about product quality":

- The product broke → yes
- Shipping damaged the product → ?
- The product is fine but not as described → ?
- The customer complains about quality *and* about shipping → ?
- The customer says the quality is fine but overpriced → ?

--

<br>

Every one of those needs an answer in the prompt. If it is not in the prompt, the model
picks, and it may not pick consistently.

---

# A prompt skeleton

.codesm[
```
You are annotating customer reviews for a research study.

CONSTRUCT: A review is coded 1 if it expresses dissatisfaction with the
physical quality, durability, or functioning of the product itself.

INCLUDE: defects, breakage, malfunction, materials wearing out, the
product not working as designed.

EXCLUDE: shipping damage, delivery delays, seller communication, price
complaints, or dissatisfaction with the product's design or aesthetics
when the product functions as intended.

EDGE CASES:
- Damage occurring in transit -> 0 (shipping, not product quality)
- Product functions but is not as described -> 0
- Both quality and non-quality complaints present -> 1
- Quality praised but price criticized -> 0

Return only JSON matching the schema. No other text.
```
]

---

# Structured outputs

Free text is not a variable. You need something parseable, and you need it to be
parseable **every time**.

--

<br>

Modern APIs will enforce a JSON schema. Use it.

--

<br>

Do not parse prose with regex. It will work on most cases, and the failures will not be
random.

---

# What to put in the schema

```json
{
  "label": 0,
  "confidence": "high",
  "evidence_span": "the hinge snapped after two weeks",
  "reasoning": "explicit report of physical failure of the product"
}
```

--

<br>

You want more than the label:

- **evidence span** so you can spot-check without rereading the document
- **reasoning** because it is diagnostic when things go wrong
- **confidence** because it is useful for stratifying your validation sample

--

<br>

.small90[Careful: a self-reported confidence is not a calibrated probability. Treat it as
a heuristic for sampling, not as a posterior.]

---

# An annotation call in R

```{r eval=FALSE}
library(httr2)
library(jsonlite)

schema <- list(
  type = "object",
  properties = list(
    label         = list(type = "integer", enum = list(0L, 1L)),
    confidence    = list(type = "string", enum = list("low","medium","high")),
    evidence_span = list(type = "string"),
    reasoning     = list(type = "string")
  ),
  required = list("label", "confidence", "evidence_span", "reasoning")
)
```

---

# An annotation call in R

```{r eval=FALSE}
annotate <- function(text, prompt, model = "claude-sonnet-4-5-20250929") {
  request("https://api.anthropic.com/v1/messages") |>
    req_headers(`x-api-key` = Sys.getenv("ANTHROPIC_API_KEY"),
                `anthropic-version` = "2023-06-01") |>
    req_body_json(list(
      model       = model,
      max_tokens  = 500,
      temperature = 0,
      system      = prompt,
      messages    = list(list(role = "user", content = text)),
      tools       = list(list(name = "record", input_schema = schema)),
      tool_choice = list(type = "tool", name = "record")
    )) |>
    req_perform() |>
    resp_body_json()
}
```

---

# The same thing in Python

```{r eval=FALSE, engine='python'}
import os, json, anthropic

client = anthropic.Anthropic(api_key=os.environ["ANTHROPIC_API_KEY"])

SCHEMA = {
    "type": "object",
    "properties": {
        "label":         {"type": "integer", "enum": [0, 1]},
        "confidence":    {"type": "string", "enum": ["low","medium","high"]},
        "evidence_span": {"type": "string"},
        "reasoning":     {"type": "string"},
    },
    "required": ["label","confidence","evidence_span","reasoning"],
}
```

---

# The same thing in Python

```{r eval=FALSE, engine='python'}
def annotate(text, prompt, model="claude-sonnet-4-5-20250929"):
    r = client.messages.create(
        model=model,
        max_tokens=500,
        temperature=0,
        system=prompt,
        messages=[{"role": "user", "content": text}],
        tools=[{"name": "record", "input_schema": SCHEMA}],
        tool_choice={"type": "tool", "name": "record"},
    )
    return r   # return the FULL response object, not r.content[0].input["label"]
```

--

.small90[Note the comment on the last line. We'll come back to it.]

---

# Temperature = 0

Set it to zero for annotation.

--

<br>

Temperature controls how much randomness is injected into token selection. Above zero,
the same document gets different labels on different calls.

--

<br>

A measurement instrument that returns different values on repeat measurement of
the same object is a broken instrument.

--

<br>

You would not accept a scale that reported a different weight each time you stepped on
it.

---

# A caveat

Temperature 0 gets you *near*-determinism, not determinism.

--

<br>

Floating-point non-associativity, batching, and infrastructure changes on the provider's
side mean identical calls can occasionally differ.

--

<br>

<br>

Which is one more reason to **save the raw outputs**. If your labels shift, you want to be
able to prove whether it was you or them.

---

# Zero-shot, few-shot, fine-tuning

**Zero-shot.** Definition only, no examples. Start here. It is the default and it is often
enough.

--

**Few-shot.** Include 5–20 labeled examples in the prompt. Use when the construct is hard
to define but easy to demonstrate, or when zero-shot output is inconsistent on edge cases.

--

**Fine-tuning.** Train on hundreds or thousands of labels. Use when you have that many
labels already, the task is a stable classification, and you are running millions of
documents.

---

# Choosing among them

.small90[
| | Labels needed | Cost per doc | When |
|:--|:--|:--|:--|
| Zero-shot | 0 | high | default; construct is definable |
| Few-shot | 10–50 | high | edge cases need demonstrating |
| Fine-tune | 1,000+ | low | huge corpus, stable target |
]

--

<br>

For a seven-week project: zero-shot, with few-shot as the fix if validation disappoints.

--

<br>

Warning: your few-shot examples must not come from your validation sample.

--

That is training on the test set.

---

# Cost control: batching

Most providers offer a batch API: submit many requests, get results within 24 hours, pay
roughly half.

--

<br>

An annotation run is not interactive. It can run overnight.

--

<br>

If your job does not need to finish in the next hour, batch it.

---

# Cost control: caching

If your prompt is long (full instructions with edge cases run 1,500 tokens), you are paying
for those tokens on every one of 200,000 calls.

--

<br>

Prompt caching lets the provider reuse the fixed prefix. Large savings when the
instruction block dwarfs the document.

--

<br>

Structure your call so the **stable part comes first** and the document last. That is what
makes caching possible.

---

# Rate limits

You will hit them. Plan for it before you start.

--

<br>

- Exponential backoff with jitter on 429s
- A concurrency cap you set explicitly
- Checkpoint after every batch, so a crash costs minutes and not hours
- Idempotent writes keyed on document ID, so restarting does not duplicate

--

<br>

.small90[Run this on one of the course servers inside `tmux`. A laptop will sleep.]

---

# Refusals and malformed output

Some fraction of calls will come back wrong. Refusals, truncations, schema violations,
timeouts.

--

<br>

The easy handling is `tryCatch` returning `NA` and moving on.

--

<br>

Do not drop them silently.

--

<br>

Refusals are not random. A model that refuses to classify is telling you something about
that document. It is likely sensitive, ambiguous, or unusual, and those are exactly the
documents your estimate is sensitive to.

---

# Handling failures

```{r eval=FALSE}
result <- tryCatch(
  annotate(doc$text, prompt),
  error = function(e) list(status = "error", message = conditionMessage(e))
)

# write EVERY outcome to disk, keyed by doc id
writeLines(toJSON(list(
  doc_id     = doc$id,
  model      = model_string,
  prompt_sha = prompt_hash,
  run_date   = Sys.time(),
  raw        = result
), auto_unbox = TRUE), con = out_con)
```

---

# Handling failures

Then, separately, report:

- how many calls failed
- what fraction of documents ended with no label
- whether failures are correlated with anything you care about

--

That last bullet goes in your paper.

---

# Store the raw output

Same rule as Lecture 1.

--

<br>

Store the **full API response**. Not `label`. The whole object: reasoning, evidence span,
confidence, model string, timestamp, token counts.

--

<br>

Three reasons:

--

- You will want to recode the construct. Most projects do.
- A referee will ask for a different threshold or a subgroup breakdown.
- The model will be deprecated, and the raw outputs will be all that survives.

--

<br>

Re-running 200k calls because you saved only the label is avoidable.

---

# The pipeline, assembled

```
prompts/
  quality-complaint-v3.md        # versioned, committed
code/
  03-measure.R                   # reads prompt, writes raw JSONL
  03b-parse.R                    # raw JSONL -> tidy labels
data/
  raw/llm/quality-v3-2026-03-14.jsonl   # never edited
  clean/labels.parquet
```

--

<br>

Two scripts, not one. Calling the API and parsing its output are different jobs with
different failure modes.

--

<br>

If they are the same script, every parsing bug costs you another $400 of API calls.

---
class: inverse, center, middle
name: validation

# Designing a validation study

---

# Not specific to AI

Before the rules: validation is not something LLMs invented a need for.

--

<br>

You owe a validation study for **any** variable that was measured rather than
observed.

--

<br>

Survey instruments get validated. Scales get validated. Hand-coding gets an
inter-coder reliability statistic and a reviewer who asks for it.

--

An LLM-coded variable is the same object. The norm already exists. It just has not
been applied to this yet.

---

# Gold-standard labels

You must have human-labeled gold-standard data.

--

<br>

Next week you will see the arithmetic reason why: without it, none of the corrections
apply, and your estimates are uninterpretable.

--

<br>

"The model is obviously right," "I read fifty and they looked fine," and "the Gilardi
paper validated it" do not count.

--

<br>

What counts is a designed, documented, human-labeled sample.

---

# Why "I read fifty and they looked fine" fails

Three separate reasons, each sufficient.

--

**1. You read them non-randomly.** You scrolled, and scrolling is not sampling.

--

**2. You read them knowing the model's answer.** That is not an independent label, that
is confirmation.

--

**3. Fifty is not enough to estimate anything.** If your construct is 10% prevalent, you
saw five positives.

---

# Who labels

**You.** Slow, expensive in your time, and you must do some of it anyway.

--

**RAs.** Cheap, trainable, need real written instructions and real supervision.

--

**Crowdworkers.** Fast and cheap at scale. Quality varies enormously and requires
attention checks.

--

**Domain experts.** Necessary when the construct requires expertise, clinical language,
legal text, technical specifications. Expensive and slow.

---

# You should personally label some

Label 100 documents yourself. By hand. Before you write the final prompt.

--

<br>

It is how you find out your construct is ambiguous.

--

<br>

You will hit document 23 and realize you do not know whether it counts. That moment
matters more than any prompt engineering: it means the *definition* is incomplete, and no
prompt fixes a definition.

---

# Humans first

Before you compare the model to humans, compare the humans to each other.

--

<br>

Two coders, same documents, independently, blind to each other and to the model.

--

<br>

If humans do not agree with each other, the construct is ill-defined, and the model cannot fix that.

---

# Inter-rater reliability

**Cohen's $\kappa$**, two raters, categorical labels.

--

Why not just report *percent agreement*? Because if 95% of documents are negative, two
coders who both label everything negative agree 95% of the time and have measured
nothing.

--

$\kappa$ subtracts off the agreement you would expect **by chance** given how often each
coder says yes.

--

$\kappa = 0$ means "no better than guessing." $\kappa = 1$ is perfect.

---

# Computing it

```r
library(irr)

val <- merge(gold, llm_out, by = "doc_id")
stopifnot(nrow(val) == nrow(gold))    # the model labelled every gold doc

kappa2(val[, c("human_label", "llm_label")])
```

--

```
 Cohen's Kappa for 2 Raters (Weights: unweighted)
 Subjects = 150
   Raters = 2
    Kappa = 0.681
```

--

<br>

.small90[`DescTools::CohenKappa(table(val$human_label, val$llm_label), conf.level = 0.95)`
gives you a confidence interval, which you should report.]

---

# What that example shows

Accuracy on that same sample: **87%**.

--

$\kappa$: **0.68**.

--

<br>

Same labels, two numbers far apart. $\kappa$ is the one to report.

--

<br>

The gap is chance agreement. The construct is present in a quarter of documents, so a
lot of the 87% is two raters both saying "no" to easy negatives.

---

# Three ways to get this wrong

**1. You never ran the model on the gold documents.** You annotated 5,000 for analysis
and 150 by hand. But the 150 must go through the model too, same prompt, same version.

--

**2. The merge silently dropped rows.** Refusals and malformed output vanish, and you
compute $\kappa$ on the easy subset. Assert the row count.

--

**3. The factor levels do not match.** `c("yes","no")` against `c(1,0)` gives you four
categories and a meaningless number.

---

# The other one you will see

**Krippendorff's $\alpha$**, same idea, more general: any number of raters, missing
data, ordinal or interval scales.

--

<br>

Use $\kappa$ for two coders and a binary label, which is your case. Use $\alpha$ if you
have three coders, or an ordinal scale, or gaps in the coding.

--

<br>

Rough conventions (and they are only conventions):

.small90[
| $\kappa$ | Reading |
|:--|:--|
| < 0.40 | your construct is not defined |
| 0.40–0.60 | weak; revise the codebook |
| 0.60–0.80 | acceptable for most published work |
| > 0.80 | strong |
]

---

# The ceiling argument

Suppose your two human coders agree at $\kappa$ = 0.60.

--

<br>

Now your LLM matches the gold standard at 0.60.

--

<br>

The LLM is at the ceiling: it agrees with the gold standard as well as another human does.

--

<br>

Reporting "our LLM only achieves 0.60" without the human-human benchmark makes your
instrument look bad when it is at the limit of the construct.

--

Reporting it *with* the benchmark shows the instrument is at the limit of the construct.

---

# What low human agreement means

It does not mean "get better RAs."

--

<br>

It means one of:

- Your construct definition is vague
- Your construct is genuinely multidimensional and you are forcing it to one dimension
- The documents do not contain enough information to determine the label

--

<br>

All three are fixed upstream, in the definition, not downstream in the prompt.

--

<br>

This is why you iterate the codebook on human labels *before* you touch the model.

---

# How many to label

It depends on what you plan to do downstream.

--

<br>

And what you plan to do downstream is **next week's topic**.

--

<br>

<br>

Working guidance now. The argument behind the number is next week.

---

# Working guidance

Hundreds, not dozens.

--

<br>

- **300** is a floor for a single binary construct
- **500–1,000** is where most credible applications sit
- More if you need accuracy *within* subgroups
- More if the positive class is rare

--

<br>

The rare-class point is the one most often missed.

---

# Scale with rarity

Your construct occurs in 2% of documents.

--

<br>

You draw a simple random sample of 500. You get **10 positives**.

--

<br>

Your estimate of how many positives the model *catches* rests on 10 cases. Its standard
error is large: you cannot distinguish recall of 0.5 from recall of 0.9.

--

<br>

Ten cases cannot tell you anything about the positive class.

--

<br>

The fix is stratification, which is the next topic.

---
# What cannot be fixed later

Most mistakes in this lecture can be fixed later. A bad prompt can be rewritten, a wrong
metric recomputed, too few labels topped up.

--

<br>

Sample the validation set the wrong way and it cannot be fixed by labeling more.

--

<br>

Unrecoverable in a specific sense: the statistical corrections you will need do not apply
and cannot be made to apply after the fact.

---

# Notation, once

Three symbols for the rest of the lecture:

--

.small90[
| | |
|:--|:--|
| $n$ | the size of your **corpus**, every document you will analyze |
| $m$ | the size of your **gold-standard sample**: the ones you label by hand |
| $\pi_i$ | the probability that document $i$ ended up in that sample |
]

--

<br>

$m$ is always much smaller than $n$. That is the point: 500 hand labels standing
in for 200,000 documents.

---

# The requirement

For every document $i$ in your corpus, there is a probability $\pi_i$ that it entered your
validation sample.

--

<br>

.large2[That probability must be **known** and **controlled by you**.]

--

<br>

Known: you can write it down as a number.

Controlled: it came from a design decision, not from what you happened to do.

---

# Sampling schemes that work

**Simple random sample.** $\pi_i = m/n$ for everyone. Draw with a seed, save the indices.
Always defensible.

--

**Stratified sample.** Partition the corpus into strata. Sample within each with known
probability $\pi_s$.

--

<br>

You are *allowed* to oversample. Oversample rare positives. Oversample the treated group.
Oversample cases where the model reports low confidence.

--

<br>

You just have to know the probabilities and record them.

--

Survey weights work the same way.

---

# Sampling schemes that do not work

- The first 500 rows of the file
- The 500 you happened to read while exploring
- The ones with the most interesting text
- The ones where the model seemed uncertain (unless you recorded $\pi$)
- The ones you labeled last June, before you thought about this
- "I kept labeling until the accuracy number looked good"

--

<br>

Every one of these produces a set of human labels. None of them produces a *sample*.

---

# Why the probabilities matter

Next week you will learn that every correction method depends on knowing
$\pi_i$.

--

<br>

DSL, prediction-powered inference, the econometrics versions, all of them.

--

<br>

They differ in their assumptions about the model, the outcome, and the estimator. They
agree on exactly one thing: you must be able to state the probability with which each
document entered your gold-standard sample.

---

# A common failure

March: you annotate 50,000 documents.

April: you run the analysis.

May: you present, and someone asks about measurement error.

June: you hand-label 500 documents "to validate."

--

<br>

Those 500 labels cannot fix the estimates.

--

<br>

They can produce an accuracy number. They cannot produce a correction, because you drew
them in June, after seeing your results, with no recorded $\pi$.

---

# Why this is a Week 4 milestone

Your validation sample must be **designed** now, before you run the model on the full
corpus, before you see any downstream result.

--

<br>

The same logic as a pre-analysis plan, or as not choosing an instrument after seeing the
first stage.

--

<br>

This is a **design** decision, not an **analysis** decision.

--

<br>

It is a Week 4 milestone because it cannot be done in Week 6.

---

# Concretely, this week

1. Decide $m$. Write the number down and justify it.

--

2. Decide the scheme. Simple random sample or stratified. If stratified, define the strata and the
   within-stratum probabilities.

--

3. Draw the sample. **Save the seed. Save the indices. Commit both.**

--

4. Record $\pi_i$ as a column in your data.

--

5. Plan to label blind to model output.

--

6. Plan to run the LLM on the validation documents too, and keep those predictions.

--

<br>

.small90[Step 6 is often skipped. You need both the truth and the prediction on the same
units.]

---

# Drawing it, in code

```{r eval=FALSE}
set.seed(20260314)          # commit this number

# stratified: oversample the rare positive class as flagged by a cheap pre-screen
corpus <- corpus |>
  mutate(stratum = case_when(
    prescreen_flag == 1 ~ "likely_positive",
    TRUE                ~ "likely_negative"
  ))

# YOU choose these. Write them in the paper.
target <- c(likely_positive = 300, likely_negative = 300)

val <- corpus |>
  group_by(stratum) |>
  mutate(N_s = n(), pi = target[stratum] / N_s) |>   # known sampling probability
  slice_sample(n = target[cur_group()$stratum]) |>
  ungroup()

write_parquet(val, "data/clean/validation-sample.parquet")   # includes pi
```

---

# Drawing it, in code

The `pi` column is the one the correction depends on, and it takes four lines of code to
create.

---

# Metrics: accuracy is not enough

One number is not enough.

--

<br>

Accuracy = fraction of documents where the model and the gold standard agree.

--

<br>

Under class imbalance, accuracy alone says little.

---

# The demonstration

Your construct has 1% **prevalence**, that is, it is actually present in 1% of
documents. 10,000 documents, 100 true positives.

--

<br>

I write a classifier. Here it is:

```{r eval=FALSE}
predict_label <- function(text) 0
```

--

<br>

Accuracy: 99%.

--

<br>

It has not identified a single case of the thing you are studying.

---

# Report the confusion matrix

.small90[
| | Gold = 1 | Gold = 0 |
|:--|:--|:--|
| **Model = 1** | TP | FP |
| **Model = 0** | FN | TN |
]

--

<br>

TP = true positives, FP = false positives, FN = false negatives.

--

**Precision** = TP / (TP + FP). Of the documents you flagged, what share were right?
*How often you are right when you say yes.*

--

**Recall** = TP / (TP + FN). Of the true positives, what share did you catch?
*How much of the real thing you found.*

**F1** = harmonic mean of the two. One number when you need one number. But report the
confusion matrix anyway.

--

Always report the 2×2.

---

# Why the 2×2 and not the summary

Precision and recall summarize the matrix. The matrix contains information they discard.

--

<br>

And next week you will find out that **false positives and false negatives affect your
downstream coefficient differently**, asymmetrically, in ways that depend on your design.

--

<br>

An F1 of 0.80 built from high precision and low recall implies a different bias than the
same F1 built the other way around.

--

<br>

<br>

You cannot reconstruct the matrix from F1. Report the matrix.

---

# The precision-recall tradeoff is yours to set

If your model outputs a probability or a confidence, you choose a threshold.

--

<br>

That choice trades precision against recall. It is a substantive decision, not a technical
one.

--

<br>

- Screening for a rare event where you will hand-check the positives? Favor recall.
- Constructing a treatment indicator where false positives contaminate the treated group?
  Favor precision.

--

<br>

Choose it before you see the downstream result, and say what you chose.

--

The reason comes later in the lecture.

---

# Validate on the population you will analyze

Your corpus is 300,000 Amazon reviews, 2019–2025, across 40 categories.

--

<br>

You validate on 500 reviews from 2019, electronics only, because that's the subset you had
loaded.

--

<br>

Your validation number describes an instrument you are not using.

---

# Distribution shift

The model's error rate is not a constant. It varies with:

- product category
- document length
- time period (language drifts; so do products)
- reviewer type
- language and register

--

<br>

If your validation sample differs from your analysis sample on any of these, your
measured accuracy does not transfer.

--

<br>

It fails silently: the number still prints.

---

# The practical rule

.large2[Draw the validation sample from the **exact frame you will analyze**, at the
**same time** you draw the analysis sample.]

--

<br>

Same corpus. Same filters. Same date range. Same everything.

--

<br>

If you later restrict your analysis sample, you must be able to restrict the validation
sample the same way, which means keeping the identifiers.

---

# Report accuracy by subgroup

One overall number hides the variation that matters.

--

<br>

Report the confusion matrix **within** the dimensions that structure your design:

- by treatment status
- by time period, if you have a before/after
- by category, if you pool across categories
- by any subgroup you will report a heterogeneous effect for

--

<br>

If accuracy differs across your treatment groups, you have a problem, and it is
the problem next week is about.

---

# The validation study, summarized

1. Design the sample **before** running the model on the corpus
2. Known, recorded $\pi_i$ for every document
3. Draw from the analysis frame, not a convenient subsample
4. Humans first: two coders, $\kappa$ or $\alpha$, blind
5. Hundreds of labels, stratified if the class is rare
6. Run the LLM on the same units and keep both
7. Confusion matrix, overall and by subgroup

--

<br>

Seven items, about two hours of thinking.

---
class: inverse, center, middle
name: sensitivity

# Prompt sensitivity

---

# A question

You wrote a prompt. Accuracy against your gold standard: 0.71.

--

<br>

You are disappointed. So you rewrite it. Now: 0.76.

--

<br>

You add three few-shot examples. 0.79.

--

<br>

You try a bigger model. 0.83.

--

<br>

You stop, and report 0.83.

---

# What did you just do?

--

<br>

You ran a specification search.

--

<br>

You tried variants of an analysis and kept the one with the most favorable number.

--

<br>

That is the same activity you would be criticized for if the variants were regression
specifications instead of prompts.

---

# The evidence

Baumann, Röttger, Urman, Wendsjö, Plaza-del-Arco, Gruber & Hovy (2025), "Large Language
Model Hacking: Quantifying the Hidden Risks of Using LLMs for Text Annotation,"
arXiv:2509.08825.

--

<br>

A working paper, not yet peer-reviewed. Read it as such.

--

<br>

The scale of the exercise is large.

---

# The design

- 37 annotation tasks
- drawn from 21 **published** studies
- 18 models
- 13 million labels generated
- 2,361 hypotheses tested

--

<br>

For each published finding, they re-run the annotation across models and prompt variants,
then re-run the downstream test and ask whether the paper's **conclusion** holds.

---

# The result

**≈31%** of hypotheses reach an incorrect conclusion with frontier models.

--

<br>

With smaller models: **≈50%**.

--

Driven purely by model and prompt choice, not by the data, and not by the hypothesis.

--

<br>

The error is not in the labels but in the *conclusions*: whether the published hypothesis
is supported.

--

<br>

Between a third and a half of the time, a defensible alternative prompt or model flips
what the paper concluded.

---

# What the number means

These are published, peer-reviewed studies, each with a measurement section that reported
an accuracy figure someone found acceptable.

--

<br>

And roughly a third of them, using the best models available, have a conclusion that a
reasonable alternative prompt would reverse.

--

<br>

<br>

Prompt choice is a researcher degree of freedom, and it is currently
unreported.

---

# The analogy

.small90[
| Regression world | Prompt world |
|:--|:--|
| Try 20 specifications, report the best | Try 20 prompts, report the best |
| Choose controls after seeing $p$ | Choose the prompt after seeing accuracy |
| Choose the sample after seeing results | Choose the model after seeing results |
| Specification curve | ...nothing, yet |
]

--

<br>

The field spent twenty years building norms against the left column.

--

<br>

The right column has no norms yet, and it is where measurement now happens.

---

# One more subtlety

In the regression case, you are searching over specifications while looking at your
*result*.

--

<br>

In the prompt case, you are searching while looking at *accuracy*, which looks harmless.

--

<br>

<br>

But accuracy and your coefficient are not independent, and next week you will see
exactly how they are linked.

--

<br>

Optimizing the first can move the second in a direction you did not choose and cannot see.

---

# Defense 1: fix the prompt first

.large2[Finalize the prompt **before** you look at any downstream result.]

--

<br>

Iterating the prompt against your *validation* labels is legitimate. That is instrument
development, and it is what your gold standard is for.

--

<br>

Iterating the prompt against your *regression coefficient* is not. That is specification
search.

--

<br>

<br>

Keep the two activities separate in time: prompt first, results after.

---

# Defense 2: pre-register the measurement

If you pre-register anything, pre-register this.

--

<br>

The prompt. The model string. The temperature. The output schema. The validation sampling
scheme and $m$. The threshold.

--

<br>

<br>

The measurement instrument is the part of your design that is easiest to change
undetectably after the fact.

--

<br>

Which makes it the part where pre-registration buys the most.

---

# Defense 3: the prompt curve

Report robustness across prompts and models, the way you would report a specification
curve.

--

<br>

- 3–5 reasonable prompt variants
- 2–3 models
- Your main estimate under each

--

<br>

If the finding survives, report the curve. If it does not, you found out before a referee did.

--

<br>

<br>

.small90[It is cheap: a few dollars and an overnight run.]

---

# What a prompt curve looks like

```{r eval=FALSE}
variants <- tibble::tribble(
  ~prompt_file,                   ~model,
  "prompts/quality-v1.md",        "claude-sonnet-4-5-20250929",
  "prompts/quality-v2-terse.md",  "claude-sonnet-4-5-20250929",
  "prompts/quality-v3-fewshot.md","claude-sonnet-4-5-20250929",
  "prompts/quality-v1.md",        "claude-opus-4-8",
  "prompts/quality-v1.md",        "gpt-4o-2024-11-20"
)

results <- variants |>
  mutate(out = purrr::map2(prompt_file, model, run_full_pipeline)) |>
  tidyr::unnest_wider(out)   # accuracy, kappa, beta_hat, se
```

--

<br>

Plot $\hat\beta$ across variants. Put it in the appendix.

--

<br>

If the estimates cluster, say so in the main text. That is evidence for the paper, not a
formality.

---

# Defense 4: version-control the prompts

This is what the `prompts/` directory from Lecture 1 was for.

--

<br>

```
prompts/
  quality-complaint-v1.md
  quality-complaint-v2.md
  quality-complaint-v3-FINAL.md    # the one used in the paper
  README.md                        # what changed and why
```

--

<br>

Every variant, committed, with a commit message saying what changed.

--

<br>

The git history is the record of what you tried.

---

# Model drift

Separate problem, same category.

--

<br>

The same prompt, sent to the same model string, gives different results in six months.

--

<br>

Providers update models behind stable aliases. They deprecate versions. Inference stacks
change.

--

<br>

You cannot `renv::snapshot()` a model.

---

# What to record

**Every time**, in your output file:

- exact model string, including the date suffix
- date and time of the call
- temperature and all sampling parameters
- SHA of the prompt file
- the full raw response

--

<br>

`claude-sonnet-4-5-20250929`, accessed 2026-03-14.

Not "we used Claude."

--

<br>

.small90[Note the training cutoff too. The reason is next week's topic.]

---

# Why this matters beyond bookkeeping

Your paper will be under review for eighteen months.

--

<br>

A referee will ask you to rerun something. The model you used will be gone.

--

<br>

<br>

If you kept the raw outputs and the validation sample, you can re-run with a new
model and check whether the conclusions survive.

--

<br>

If you kept a column of zeros and ones, you cannot. You will re-annotate the corpus, get
different numbers, and have no way to tell whether the difference is the model or you.

---
class: inverse, center, middle
name: close

# Where this leaves you

---

# What you can do now

You can take a construct that has no column in any dataset and build a variable for it.

--

<br>

You can write a codebook, encode it as a prompt, enforce a schema, run it on a corpus, and
control the cost.

--

<br>

You can design a validation study, a real one, with a known sampling probability, human
inter-rater reliability, and a confusion matrix.

--

<br>

Three years ago that was months of infrastructure. Now it is a week.

---

# Suppose you did all of it

You wrote a careful prompt. You hand-labeled 500 documents, drawn as a stratified sample
with recorded $\pi_i$, before you ran anything on the full corpus.

--

<br>

You checked inter-rater reliability with an RA: $\kappa$ = 0.81.

--

<br>

Your classifier agrees with the gold standard **88%** of the time. Precision 0.86, recall
0.90. You have the full confusion matrix, overall and by treatment group.

--

<br>

You ran it on all 200,000 documents. You have your variable.

---

# Then the regression

That is a better validation study than most of what is in print.

--

<br>

<br>

You put the variable in a regression.

--

<br>

$$y_i = \alpha + \beta \widehat{D}_i + X_i'\gamma + \varepsilon_i$$

--

.small90[
$y_i$ your outcome &nbsp;·&nbsp; $\widehat{D}_i$ **the label the model gave you** &nbsp;·&nbsp;
$X_i$ controls &nbsp;·&nbsp; $\beta$ the number you care about &nbsp;·&nbsp;
$\alpha$ the intercept, not Krippendorff's
]

--

The hat on $\widehat{D}$ matters. It says this is an **estimate**, not something you
observed.

--

<br>

You get a coefficient, a standard error, and a $t$-statistic of 7.8.

---

# Next week

That regression is biased and its confidence intervals are invalid.

--

<br>

Not because you did the validation badly. You did it well.

--

<br>

It happens at 88%, at 93%, at any accuracy short of perfect.

--

<br>

<br>

And the size of the problem is not 12%.

---

# What next week needs from this week

You now know how to build an instrument and check it.

--

<br>

Next week is about what happens when you *use* it, why classification error here is not
the measurement error your econometrics sequence taught you, why "my estimates are
conservative" does not follow, and what to do instead.

--

<br>

<br>

It only works if this week's design was done.

---

# Week 4 milestone

Due before next class:

--

**1. Measurement approach specified.** What construct, from what documents, why an LLM
rather than a dictionary or a fine-tuned classifier.

--

**2. Prompt and model documented.** The prompt file committed to `prompts/`. The exact
model string. Temperature. Output schema.

--

**3. Validation sample designed**, not collected.

--

<br>

.small90[$m$, the sampling scheme, the strata if any, the probabilities $\pi_i$, who is
labeling, and how you will keep them blind. One page, in `memo/validation-design.md`,
committed.]

---

# Why "designed, not collected"

Because I want to look at your sampling scheme **before** you spend two weeks labeling
under a bad one.

--

<br>

Fixing a sampling design on paper takes ten minutes.

--

<br>

Fixing it after you have labeled 500 documents takes another 500 documents.

--

<br>

<br>

Bring the design to office hours.

---

# Reading for next week

**Egami, Hinck, Stewart & Wei (2023)**, "Using Imperfect Surrogates for Downstream
Inference," *NeurIPS* 36.

**Angelopoulos, Bates, Fannjiang, Jordan & Zrnic (2023)**, "Prediction-Powered Inference,"
*Science* 382(6671):669–674.

**Wang, Zhang & Zhang (2026)**, "Large Language Models for Market Research: A Data-Augmentation Approach," *Marketing
Science* 45(4):728–751.

--

<br>

Read the first one carefully. The other two for the framing.

--

<br>

.small90[If you read only one section of Egami et al., read the one on the sampling design.
It is today's lecture, from the other side.]

---
class: inverse, center, middle

# Questions?

### Bring your validation design to office hours before you label anything.
